---
title: "prelim_2: Bigram Logistic PCR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(tidymodels)
library(tidytext)
library(here)
# for PCA and logistic regression
library(broom)

# source preprocessing functions (adjust path if needed)
source(here("scripts/prelim_2_Wendy/Prelim_2_Preprocessing.R"))
```

## 1. Data and preprocessing

```{r load-data}
# load raw claims data
load(here("data/claims-raw.RData"))

# apply HTML parsing and cleaning
claims_clean <- claims_raw %>%
  parse_data()

# keep only rows with non-missing text and labels
claims_clean <- claims_clean %>%
  filter(!is.na(text_clean), !is.na(bclass)) %>%
  distinct(.id, .keep_all = TRUE)

claims_clean %>% glimpse()
```


## 2. Tokenization: words and bigrams

We now create two feature sets:

- a **word-level tf-idf** matrix (using `nlp_fn`)
- a **bigram-level tf-idf** matrix (using `nlp_bigram_fn`)


```{r, tokenization}
# word-level features
word_df <- claims_clean %>%
  nlp_fn()

# bigram-level features
bigram_df <- claims_clean %>%
  nlp_bigram_fn()

# align on common ids and order
ids_common <- intersect(word_df$.id, bigram_df$.id)

word_df <- word_df %>%
  filter(.id %in% ids_common) %>%
  arrange(.id)

bigram_df <- bigram_df %>%
  filter(.id %in% ids_common) %>%
  arrange(.id)

# check alignment
all(word_df$.id == bigram_df$.id)
```

## 3. Logistic principal component regression on word tokens

We:

1. Construct a tf-idf matrix from word tokens.
2. Perform PCA.
3. Use a subset of principal components as predictors in a logistic regression for `bclass`.


```{r, pcr-word}
# response variable: binary 0/1
y <- word_df$bclass %>%
  droplevels() %>%
  as.numeric() - 1

# word tf-idf matrix
X_word <- word_df %>%
  select(-.id, -bclass) %>%
  as.matrix()

# PCA on word tf-idf
pca_word <- prcomp(X_word, center = TRUE, scale. = TRUE)

# choose number of PCs to explain (e.g.) 90% of the variance
var_explained_word <- cumsum(pca_word$sdev^2) / sum(pca_word$sdev^2)
k_word <- which(var_explained_word >= 0.90)[1]

k_word
```

Our output, k_word, is 875. This number indicates that after conducting a Principal Component Analysis (PCA) on the word-level tf-idf matrix, we are able to explain 90% of total variance with just the first 875 principal components.

Next, we used these 875 principal components as predictors in a logistic regression model to estimate how well the reduced word-level features can perform in a binary classification model.

```{r, glm-training}
# add PCs from word tf_idf
PCs_idf <- pca_word$x

# create a df for PCs_idf across y
word_pcr.df <- data.frame(
  PCs_idf,
  y = y
)

# logistic regression model using our PCs
word_pcr.glm <- glm(y ~ ., family = binomial, data = word_pcr.df)

# binary classification model
# [0,1]
p_word <- predict(word_pcr.glm, type = 'response')
# if >.5 then 1, else 0
p_hat_word <- as.integer(p_word > 0.5)
# training accuracy
word_score <- mean(p_hat_word == y)
word_score
```

After conducting a Principal Component Analysis on word-level tf-idf features, the logistic regression model correctly classifies about 80.66% of the documents.

## 4. Logistic PCR with Word log-odds & Bigram PCs

```{r, logit-bigram-PCA}
# bigram tf-idf
bigram_pcs <- bigram_df %>% 
  select(-.id, -bclass) %>% 
  select(where(is.numeric)) %>% 
  # was having ram trouble here
  select(where(~ sum(. != 0) >= 5)) %>% 
  as.matrix()

# PCA on bigram mx  
pca_bigram <- prcomp(bigram_pcs, bigramcenter = TRUE, scale. = TRUE)

# aim for 90% var explained
var_explained_bigram <- cumsum(pca_bigram$sdev^2) / sum(pca_bigram$sdev^2)
k_bigram <- which(var_explained_bigram >= .90)[1]
k_bigram

```

Our output, k_bigram, is 467. This number indicates that after conducting a Principal Component Analysis (PCA) on the bigram tf-idf matrix, we are able to explain 90% of total variance with just the first 467 principal components.

Next, we implemented a logistic regression model that combines the predicted log-odds from the word-level PCR model with the 467 PCs calculated from the bigram tf-idf to determine whether incorporating bigram information improves binary classification performance.

```{r, logit-bigram.glm}
# 90% var bigram
PCs_bigram <- pca_bigram$x[, 1:k_bigram]

# log-odds from word_pcr.glm
logit_word <- predict(word_pcr.glm, type = "link")

link_df <- data.frame(
  logit_word = logit_word,
  PCs_bigram,
  y = y
)

logit_bigram.glm <- glm(y ~ ., family = binomial, data = link_df)

# binary classification model
# [0,1]
p_logbig <- predict(logit_bigram.glm, type = 'response')
# if >.5 then 1, else 0
p_hat_logbig <- as.integer(p_logbig > 0.5)
# training accuracy
logbig_score <- mean(p_hat_logbig == y)
logbig_score
```

When the predicted log-odds from the word model are combined with a Principal Components Analysis on bigram features, the resulting logistic regression model correctly classifies about 85.03% of the documents. With a higher accuracy than the word-only model, this improved performance could indicate that bigrams capture additional information about the claims status of a page.