---
title: "prelim_2: Bigram Logistic PCR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(tidymodels)
library(tidytext)

# for PCA and logistic regression
library(broom)

# source preprocessing functions (adjust path if needed)
source("scripts/prelim_2_Wendy/Prelim_2_Preprocessing.R")
```

## 1. Data and preprocessing

```{r load-data}
# load raw claims data
load("data/claims-raw.RData")

# apply HTML parsing and cleaning
claims_clean <- claims_raw %>%
  parse_data()

# keep only rows with non-missing text and labels
claims_clean <- claims_clean %>%
  filter(!is.na(text_clean), !is.na(bclass)) %>%
  distinct(.id, .keep_all = TRUE)

claims_clean %>% glimpse()
```


## 2. Tokenization: words and bigrams

We now create two feature sets:

- a **word-level tf-idf** matrix (using `nlp_fn`)
- a **bigram-level tf-idf** matrix (using `nlp_bigram_fn`)

```r
```{r tokenization}
# word-level features
word_df <- claims_clean %>%
  nlp_fn()

# bigram-level features
bigram_df <- claims_clean %>%
  nlp_bigram_fn()

# align on common ids and order
ids_common <- intersect(word_df$.id, bigram_df$.id)

word_df <- word_df %>%
  filter(.id %in% ids_common) %>%
  arrange(.id)

bigram_df <- bigram_df %>%
  filter(.id %in% ids_common) %>%
  arrange(.id)

# check alignment
all(word_df$.id == bigram_df$.id)
```

## 3. Logistic principal component regression on word tokens

We:

1. Construct a tf-idf matrix from word tokens.
2. Perform PCA.
3. Use a subset of principal components as predictors in a logistic regression for `bclass`.

```r
```{r pcr-word}
# response variable: binary 0/1
y <- word_df$bclass %>%
  droplevels() %>%
  as.numeric() - 1

# word tf-idf matrix
X_word <- word_df %>%
  select(-.id, -bclass) %>%
  as.matrix()

# PCA on word tf-idf
pca_word <- prcomp(X_word, center = TRUE, scale. = TRUE)

# choose number of PCs to explain (e.g.) 90% of the variance
var_explained_word <- cumsum(pca_word$sdev^2) / sum(pca_word$sdev^2)
k_word <- which(var_explained_word >= 0.90)[1]

k_word
```
